A greedy algorithm for the cardinality-constrained college application problem

### 2022-08-18 paper_alma/abstract.tex
intro

This paper considers a novel portfolio optimization problem called the college application problem. We show that the objective function is a nondecreasing submodular set function and provide a quadratic-time solution algorithm. We discuss the properties of the optimal solution and their implications for student welfare.

>>>>

This paper considers a novel portfolio optimization problem called the college application
problem. We show that the objective function is a nondecreasing submodular set function
and provide a quadratic-time algorithm for its maximization over a cardinality
constraint. The algorithm's validity rests on the nested structure of the optimal solution,
a property that is of independent interest. Further analysis of the optimal application strategy
helps explain perplexing findings from regression studies of college applicant behavior:
In our model, systematic differences in the risk allocations of low- and high-income students
emerge spontaneously from differences in applicants' budgets, even in the absence of risk aversion.
Also, our analysis suggests that the underwhelming effectiveness of behavioral nudges
(such as application fee waivers) designed to encourage talented low-income students to apply
to more colleges can be attributed to the fact that these interventions have typically targeted
students with small shadow costs.

### 2022-08-18 paper_alma/body.tex
intro

The college application problem can be viewed as a static variant of the Pandora's Box problem \citep{weitzman1979,boodaghiansetal2022}. In a Pandora's Box formulation of college application, the student applies to schools one by one, each time paying an application fee and observing her admissions outcome after a certain time delay. The problem is to determine an optimal \emph{stopping rule} for when Pandora should halt her college search and accept the best admissions offer she has on hand. \cite{weitzman1979} showed that the optimal policy is to stop searching when the value of the current best offer exceeds the maximum \emph{reservation price,} a statistic that represents the expected value of applying to a new college.

Arguably, the static model considered in this study is more hostile to students than the Pandora's Box problem. If, for example, an unlucky Pandora is rejected from a safety school at an early round of application, then she can compensate for the unexpected loss by pivoting to a more risk-averse application strategy. By contrast, the decisionmaker in our college application problem must commit at the outset to applying to every school in her portfolio. The admissions process used in the United States can be viewed as the concatenation of both problems: In the fall, students solve \eqref{headlineproblem} and send out a batch of applications. Then, upon observing their admissions outcomes in March, they use the Pandora strategy to pursue additional offers by applying to schools that offer rolling admissions.

>>>>

In a narrow sense, the college application problem is a static variant of the Pandora's Box problem \citep{weitzman1979,boodaghiansetal2022}. In a Pandora's Box formulation of college application, the student applies to schools one by one, each time paying an application fee and observing her admissions outcome after a certain time delay. The problem is to determine an optimal \emph{halting rule} for when Pandora should stop applying to new schools and enroll at the best school she has been admitted to so far. \cite{weitzman1979} showed that the optimal policy is to end the search when the value of the current best offer exceeds the maximum \emph{reservation price,} a statistic that represents the marginal value of an additional application.

The static model considered in this study is arguable more hostile to students than the Pandora's Box problem. If, for example, an unlucky Pandora is rejected from a safety school at an early round of application, then she can compensate for the unexpected loss by pivoting to a more risk-averse application strategy. By contrast, the decisionmaker in our college application problem must commit at the outset to applying to every school in her portfolio. The admissions process used in the United States can be viewed as the concatenation of both problems: In the fall, students solve \eqref{headlineproblem} and send out a batch of applications. Then, upon observing their admissions outcomes in March, they use the Pandora strategy to pursue additional offers by applying to schools that offer rolling admissions.
